---
title: "Getting started with cpaic"
output: rmarkdown::html_vignette
bibliography: cpaic.bib
vignette: >
  %\VignetteIndexEntry{Getting started with cpaic}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6,
                      fig.height = 4)
```

```{r setup}
library(cpaic)
```

## The problem

Network meta-analysis needs a *connected* network. When the evidence splits
into two or more sub-networks with no common comparator, the network is
**disconnected** and the treatments in different sub-networks cannot be
compared directly.

`cpaic` reconnects such a network through the *additive component* structure
of the treatments (component network meta-analysis), and then adjusts the
comparisons for between-study differences in effect modifiers using anchored
population-adjustment methods (STC, MAIC, and ML-NMR). The
result is an indirect comparison that is both connected and
population-adjusted.

## A disconnected example

The bundled data describe a binary-outcome network in two pieces:

* sub-network 1, anchored on placebo: `Placebo`, `A`, `B`;
* sub-network 2, isolated: `A+B`, `A+B+C`, `A+B+D`.

No treatment is shared between the two pieces, so the network is
disconnected. The shared components `A` and `B` bridge it.

```{r}
net <- cpaic_network(cpaic_bin_agd, ipd = cpaic_bin_ipd, sm = "OR",
                     family = "binomial", ipd_covariates = "x1",
                     inactive = "Placebo")
net
```

## Is the network bridgeable?

A disconnected network can be bridged only if the shared components make all
component effects identifiable, that is `rank(X) = number of components`.

```{r}
cpaic_connectivity(net)
```

The report confirms two sub-networks, identifies `A` and `B` as the bridging
components, and shows the component effects are identifiable.

```{r, fig.alt = "Component network coloured by sub-network"}
plot(net)
```

## Step 1: connect with component NMA

`cnma_bridge()` fits the additive component model and reconstructs the
relative effects across the gap.

```{r}
br <- cnma_bridge(net)
component_effects(br)
```

## Step 2: adjust for effect modifiers

Components `C` and `D` come from the IPD studies, whose effect modifier `x1`
is imbalanced relative to the target population (`x1 = 0`). Anchored STC fits
an outcome regression with treatment-by-`x1` interactions and reads off the
treatment effect at the target.

```{r}
fit_stc <- cstc(net, target = c(x1 = 0), effect_modifiers = "x1")
component_effects(fit_stc)
```

Anchored MAIC instead reweights each IPD study to the target population.

```{r}
fit_maic <- cmaic(net, target = c(x1 = 0), effect_modifiers = "x1",
                  n_boot = 100, seed = 1)
effective_sample_size(fit_maic)
```

Population adjustment moves the `C` and `D` effects relative to the
unadjusted (naive) bridge, while the placebo-anchored components `A` and `B`
are unchanged:

```{r}
data.frame(
  component = component_effects(br)$component,
  naive     = round(component_effects(br)$estimate, 3),
  cSTC      = round(component_effects(fit_stc)$estimate, 3),
  cMAIC     = round(component_effects(fit_maic)$estimate, 3)
)
```

## Reporting

```{r}
relative_effects(fit_stc)
additivity_test(fit_stc)
```

```{r, fig.alt = "Forest plot of relative effects versus placebo"}
forest(fit_stc)
```

## Where next

* `vignette("cpaic-methods")` covers the statistical framework in depth.
* The full mathematical foundations and a validation study are provided with
  the development sources.
