---
title: "The cpaic statistical framework"
output: rmarkdown::html_vignette
bibliography: cpaic.bib
vignette: >
  %\VignetteIndexEntry{The cpaic statistical framework}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(cpaic)
```

This vignette describes the statistical framework behind cpaic. The
companion mathematical-foundations document (shipped with the development
sources) gives the full derivations; here we summarize the model and show
how each piece maps to a function.

## Two layers

cpaic targets networks that are *disconnected*: the treatments split into
two or more sub-networks with no common comparator, so standard network
meta-analysis cannot compare across the gap. Two layers solve the two
distinct problems.

1. **Connection layer (component network meta-analysis).** Multi-component
   treatments are decomposed into additive component effects. When
   sub-networks share components, the component effects bridge the gap.
2. **Adjustment layer (population adjustment).** Where individual patient
   data (IPD) are available, each evidence edge is corrected for
   between-study imbalance in effect modifiers, using anchored STC, MAIC,
   or ML-NMR.

The output is an indirect comparison that is both connected and
population-adjusted.

## The additive component model

Let $\delta$ be the vector of observed relative effects (one per
comparison), $B$ the edge-incidence (contrast) matrix mapping comparisons
to treatments, and $C$ the treatment-by-component matrix with
$C_{tc} = 1$ if treatment $t$ contains component $c$. Treatment effects are
additive in the component effects $\beta$,
$$ \theta = C\beta, \qquad \delta = B\theta = BC\beta = X\beta, $$
with $X = BC$ the component design matrix. With inverse-variance weights
$W$, the component effects are estimated by weighted least squares
$$ \hat\beta = (X^\top W X)^{+} X^\top W d, \qquad
   \mathrm{Cov}(\hat\theta) = C (X^\top W X)^{+} C^\top, $$
where $(\cdot)^{+}$ is the Moore-Penrose inverse and $d$ the data vector
[@rucker2020cnma]. The additivity assumption is checked with a Cochran $Q$
statistic. This is implemented in `cnma_bridge()`, a wrapper around
`netmeta::discomb()`.

## Connecting a disconnected network

A disconnected network can be bridged only if the shared components make
the component effects identifiable, that is
$\mathrm{rank}(X) = K$, the number of components. `cpaic_connectivity()`
detects the sub-networks, lists the bridging components, and reports
identifiability.

```{r}
net <- cpaic_network(cpaic_bin_agd, sm = "OR", inactive = "Placebo")
cpaic_connectivity(net)
```

When identifiable, `cnma_bridge()` reconstructs the relative effects across
the gap from the component effects.

```{r}
component_effects(cnma_bridge(net))
```

## Anchored simulated treatment comparison (cSTC)

For each IPD study, `cstc()` fits an outcome regression with treatment main
effects, prognostic main effects, and treatment-by-effect-modifier
interactions, with the effect modifiers centered at a target population.
On the link scale,
$$ g\{E(y \mid \text{arm } t, x)\}
   = \mu + \beta_t + \gamma_t^\top (x - \bar x_{\text{target}}) + \dots $$
so the treatment coefficient $\beta_t$ is the population-adjusted contrast
at the target (the interaction term vanishes at $x = \bar x_{\text{target}}$).
This is the anchored generalization of regression-based standardization; it
is implemented natively because `mlumr::stc()` targets the unanchored
two-trial case.

```{r}
net_ipd <- cpaic_network(cpaic_bin_agd, ipd = cpaic_bin_ipd, sm = "OR",
                         family = "binomial", ipd_covariates = "x1",
                         inactive = "Placebo")
component_effects(cstc(net_ipd, target = c(x1 = 0), effect_modifiers = "x1"))
```

## Anchored matching-adjusted indirect comparison (cMAIC)

`cmaic()` reweights each IPD study so that its effect-modifier distribution
matches the target population, using entropy-balancing weights
$w_i = \exp(\tilde x_i^\top \alpha)$ with $\tilde x_i$ the centered effect
modifiers [@phillippo2020mlnmr]. The effective sample size is
$\mathrm{ESS} = (\sum_i w_i)^2 / \sum_i w_i^2$. The weighted within-study
contrasts, with bootstrap standard errors that propagate the weighting
uncertainty, are then combined through the component model.

```{r}
fit_maic <- cmaic(net_ipd, target = c(x1 = 0), effect_modifiers = "x1",
                  n_boot = 100, seed = 1)
effective_sample_size(fit_maic)
```

## The unification

For a single IPD edge anchored on a common comparator, combining the
adjusted contrast with the aggregate comparator contrast through the
component model is exactly a Bucher indirect comparison. The component
model generalizes this to a network: many adjusted and unadjusted contrasts
are combined simultaneously through $\hat\beta = (X^\top W X)^{+} X^\top W d$,
and the bridge supplies the contrasts that are otherwise unavailable across
the disconnection. The key assumption is that component effects (and their
effect-modifier interactions) are constant across sub-networks
[@rucker2021disconnected].

## Component-additive ML-NMR

`cmlnmr()` places the component structure inside multilevel network
meta-regression: the relative effect of an arm is $C\beta$ rather than a
free per-treatment parameter, and aggregate arms are fitted by integrating
the individual model over each study's covariate distribution. Disconnected
sub-networks share the component parameters, so the network is connected by
construction. All four outcome families are supported; survival uses a
proportional-hazards model with a flexible baseline (piecewise-exponential
by default, or a smooth M-spline), set through `cut_points` and `baseline`.
Aggregate covariates are integrated with a Gaussian copula whose
correlation is estimated from the individual patient data. A component
whose effect-modifier interaction is informed only by aggregate data is
weakly identified and relies on the regression prior.

```{r, eval = FALSE}
# Requires cmdstanr; see ?cmlnmr.
fit <- cmlnmr(ipd, agd, effect_modifiers = "x1", inactive = "Placebo",
              family = "binomial")
component_effects(fit)
```

## Assumptions and caveats

* **Additivity** of component effects (test with `additivity_test()`; add
  interaction terms if violated).
* **Identifiability**: a disconnected network is bridgeable only when
  `cpaic_connectivity()` reports `identifiable = TRUE`.
* **Cross-population transportability** of effect modifiers (the standard
  population-adjustment assumption), here extended to constancy of
  component effects across sub-networks.
* **Non-collapsibility**: MAIC targets a marginal effect and STC a
  conditional effect, so for noncollapsible measures they answer slightly
  different questions.

## References
