---
title: "The cpaic statistical framework"
output: rmarkdown::html_vignette
bibliography: cpaic.bib
vignette: >
  %\VignetteIndexEntry{The cpaic statistical framework}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(cpaic)
```

> **Research use only.** cpaic is experimental. Its methodology and
> implementation have not been validated for clinical, regulatory,
> reimbursement, or other decision use.

This vignette describes the statistical framework behind cpaic. The
companion mathematical-foundations document (shipped with the development
sources) gives the full derivations; here we summarize the model and show
how each piece maps to a function.

## Two layers

cpaic targets networks that are *disconnected*: the treatments split into
two or more sub-networks with no common comparator, so standard network
meta-analysis cannot compare across the gap. Two layers solve the two
distinct problems.

1. **Connection layer (component network meta-analysis).** Multi-component
   treatments are decomposed into additive component effects. When
   sub-networks share components, the component effects bridge the gap.
2. **Adjustment layer (population adjustment).** Where individual patient
   data (IPD) are available, each evidence edge is corrected for
   between-study imbalance in effect modifiers, using anchored STC, MAIC,
   or ML-NMR.

The component model can identify a cross-gap contrast, but scientific
transportability is an additional assumption. A two-stage result is not a
coherent single-target population adjustment when target-specific IPD
contrasts are mixed with retained AgD contrasts from their original study
populations.

## The additive component model

Let $\delta$ be the vector of observed relative effects (one per
comparison), $B$ the edge-incidence (contrast) matrix mapping comparisons
to treatments, and $C$ the treatment-by-component matrix with
$C_{tc} = 1$ if treatment $t$ contains component $c$. Treatment effects are
additive in the component effects $\beta$,
$$ \theta = C\beta, \qquad \delta = B\theta = BC\beta = X\beta, $$
with $X = BC$ the component design matrix. With inverse-variance weights
$W$, the component effects are estimated by weighted least squares
$$ \hat\beta = (X^\top W X)^{+} X^\top W d, \qquad
   \mathrm{Cov}(\hat\theta) = C (X^\top W X)^{+} C^\top, $$
where $(\cdot)^{+}$ is the Moore-Penrose inverse and $d$ the data vector
[@rucker2020cnma]. The additivity assumption is checked with a Cochran $Q$
statistic. This is implemented in `cnma_bridge()`, a wrapper around
`netmeta::discomb()`.

## Connecting a disconnected network

A full set of component effects is identifiable when
$\mathrm{rank}(X) = K$, the number of components. Even in a rank-deficient
design, a particular contrast can be identifiable when its contrast vector
lies in the row space of $X$. `cpaic_connectivity()` detects the sub-networks
and reports the component-design diagnostics; `estimable_effects()` checks the
requested contrasts.

```{r}
net <- cpaic_network(cpaic_bin_agd, sm = "OR", inactive = "Placebo")
cpaic_connectivity(net)
```

When identifiable, `cnma_bridge()` reconstructs the relative effects across
the gap from the component effects.

```{r}
component_effects(cnma_bridge(net))
```

## Anchored simulated treatment comparison (cSTC)

For each IPD study, `cstc()` fits an outcome regression with treatment main
effects, prognostic main effects, and treatment-by-effect-modifier
interactions, with the effect modifiers centered at supplied target covariate
means.
On the link scale,
$$ g\{E(y \mid \text{arm } t, x)\}
   = \mu + \beta_t + \gamma_t^\top (x - \bar x_{\text{target}}) + \dots $$
so the treatment coefficient $\beta_t$ is the conditional link-scale contrast
at the supplied target covariate means (the interaction term vanishes at
$x = \bar x_{\text{target}}$). Because this contrast is linear in the
effect modifiers on the link scale, it is also the average conditional
link-scale contrast at those means. It is not an outcome-scale marginal
standardization. The model is implemented natively.

```{r}
net_ipd <- cpaic_network(cpaic_bin_agd, ipd = cpaic_bin_ipd, sm = "OR",
                         family = "binomial", ipd_covariates = "x1",
                         inactive = "Placebo")
component_effects(cstc(net_ipd, target = c(x1 = 0),
                       effect_modifiers = "x1",
                       allow_experimental_bridge = TRUE))
```

## Anchored matching-adjusted indirect comparison (cMAIC)

`cmaic()` reweights each IPD study so that its effect-modifier moments match
the supplied target moments, using entropy-balancing weights
$w_i = \exp(\tilde x_i^\top \alpha)$ with $\tilde x_i$ the centered effect
modifiers [@phillippo2020mlnmr]. The effective sample size is
$\mathrm{ESS} = (\sum_i w_i)^2 / \sum_i w_i^2$. The weighted within-study
contrasts, with bootstrap standard errors that propagate the weighting
uncertainty, are then combined through the component model.

That last step is an additional methodology constraint. A marginal weighted
contrast is not generally additive across treatment components on a nonlinear
link scale. Also, if AgD-only edges remain, those contrasts still refer to
their original study populations. cMAIC therefore stops by default outside the
Gaussian identity-link case and whenever retained AgD edges would be mixed in.
The experimental override does not repair either estimand mismatch.

```{r}
fit_maic <- cmaic(net_ipd, target = c(x1 = 0), effect_modifiers = "x1",
                  n_boot = 100, seed = 1,
                  allow_experimental_bridge = TRUE)
effective_sample_size(fit_maic)
```

## The unification

For a single anchored comparison, the contrast algebra resembles a Bucher
indirect comparison. A network bridge combines many adjusted and unadjusted
contrasts through $\hat\beta = (X^\top W X)^{+} X^\top W d$. That algebra does
not establish that every input contrast shares the same target estimand. It
also does not test the required constancy of component effects and
effect-modifier interactions across disconnected sub-networks
[@rucker2021disconnected].

## Component-additive ML-NMR

`cmlnmr()` places the component structure inside multilevel network
meta-regression: the relative effect of an arm is $C\beta$ rather than a
free per-treatment parameter, and aggregate arms are fitted by integrating
the individual model over each study's covariate distribution. Disconnected
sub-networks share the component parameters, so the network is connected by
construction. All four outcome families are supported; survival uses a
proportional-hazards model with a flexible baseline (piecewise-exponential
by default, or a smooth M-spline), set through `cut_points` and `baseline`.
Its analytic row-level likelihood handles events, right/left/interval
censoring, and delayed entry. Aggregate survival arms require reconstructed
event and censoring rows; event counts plus person-time are rejected. The
finite quasi-Monte Carlo covariate integration is approximate. Restricted mean
survival time is not implemented.
Aggregate covariates are integrated with a Gaussian copula whose
correlation is estimated from the individual patient data. A component
whose effect-modifier interaction is informed only by aggregate data is
weakly identified and relies on the regression prior.

Reporting methods evaluate
$(C_t-C_u)'(\beta + \Gamma x)$ at a one-row vector of supplied covariate
means. Since this expression is linear in $x$, it is an average conditional
link-scale effect at those means. It is not a marginally standardized odds,
rate, or hazard ratio. The aggregate likelihood integrates outcomes over
covariate distributions for model fitting, but that integration does not turn
the reporting function into marginal standardization.

```{r, eval = FALSE}
# Fits with rstan by default; backend = "cmdstanr" is the alternative.
fit <- cmlnmr(ipd, agd, effect_modifiers = "x1", inactive = "Placebo",
              family = "binomial")
component_effects(fit)
```

## Assumptions and caveats

* **Additivity** of component effects. `additivity_test()` can diagnose lack of
  fit where evidence exists, but it cannot validate an unobserved cross-gap
  bridge.
* **Identifiability**: a disconnected network is bridgeable only when
  `cpaic_connectivity()` reports `identifiable = TRUE`.
* **Cross-population transportability** of effect modifiers (the standard
  population-adjustment assumption), here extended to constancy of
  component effects across sub-networks.
* **Two-stage estimand mixing**: retained AgD contrasts remain tied to their
  study populations. The default gate stops this mixed bridge unless the
  caller explicitly opts into research-only behavior.
* **Nonlinearity and non-collapsibility**: MAIC targets a marginal weighted
  contrast and STC a conditional link-scale contrast. Marginal contrasts are
  not generally component-additive on nonlinear links.
* **cML-NMR reporting**: target summaries are average conditional link-scale
  effects at covariate means, not marginally standardized effects.

## References
