---
title: "The cpaic statistical framework"
output: rmarkdown::html_vignette
bibliography: cpaic.bib
vignette: >
  %\VignetteIndexEntry{The cpaic statistical framework}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(cpaic)
```

> **Research use only.** cpaic is experimental. Its methodology and
> implementation have not been validated for clinical, regulatory,
> reimbursement, or other decision use.

This vignette describes the statistical framework behind cpaic. The
companion mathematical-foundations document (shipped with the development
sources) gives the full derivations; here we summarize the model and show
how each piece maps to a function.

## Two layers

cpaic targets networks that are *disconnected*: the treatments split into
two or more sub-networks with no common comparator, so standard network
meta-analysis cannot compare across the gap. Two layers solve the two
distinct problems.

1. **Connection layer (component network meta-analysis).** Multi-component
   treatments are decomposed into additive component effects. When
   sub-networks share components, the component effects bridge the gap.
2. **Adjustment layer (population adjustment).** Where individual patient
   data (IPD) are available, each evidence edge is corrected for
   between-study imbalance in effect modifiers, using anchored STC, MAIC,
   or ML-NMR.

The component model can identify a cross-gap contrast, but scientific
transportability is an additional assumption. A two-stage result is not a
coherent single-target population adjustment when target-specific IPD
contrasts are mixed with retained AgD contrasts from their original study
populations.

## The additive component model

Let $\delta$ be the vector of observed relative effects (one per
comparison), $B$ the edge-incidence (contrast) matrix mapping comparisons
to treatments, and $C$ the treatment-by-component matrix with
$C_{tc} = 1$ if treatment $t$ contains component $c$. Treatment effects are
additive in the component effects $\beta$,
$$ \theta = C\beta, \qquad \delta = B\theta = BC\beta = X\beta, $$
with $X = BC$ the component design matrix. With inverse-variance weights
$W$, the component effects are estimated by weighted least squares
$$ \hat\beta = (X^\top W X)^{+} X^\top W d, \qquad
   \mathrm{Cov}(\hat\theta) = C (X^\top W X)^{+} C^\top, $$
where $(\cdot)^{+}$ is the Moore-Penrose inverse and $d$ the data vector
[@rucker2020cnma]. The additivity assumption is checked with a Cochran $Q$
statistic. This is implemented in `cnma_bridge()`, a wrapper around
`netmeta::discomb()`.

## Connecting a disconnected network

A full set of component effects is identifiable when
$\mathrm{rank}(X) = K$, the number of components. Even in a rank-deficient
design, a particular contrast can be identifiable when its contrast vector
lies in the row space of $X$. `cpaic_connectivity()` detects the sub-networks
and reports the component-design diagnostics; `estimable_effects()` checks the
requested contrasts.

```{r}
net <- cpaic_network(cpaic_bin_agd, sm = "OR", inactive = "Placebo")
cpaic_connectivity(net)
```

When identifiable, `cnma_bridge()` reconstructs the relative effects across
the gap from the component effects.

```{r}
component_effects(cnma_bridge(net))
```

## Anchored simulated treatment comparison (cSTC)

For each IPD study, `cstc()` fits an outcome regression with treatment main
effects, prognostic main effects, and treatment-by-effect-modifier
interactions, with the effect modifiers centered at supplied target covariate
means.
On the link scale,
$$ g\{E(y \mid \text{arm } t, x)\}
   = \mu + \beta_t + \gamma_t^\top (x - \bar x_{\text{target}}) + \dots $$
so the treatment coefficient $\beta_t$ is the conditional link-scale contrast
at the supplied target covariate means (the interaction term vanishes at
$x = \bar x_{\text{target}}$). Because this contrast is linear in the
effect modifiers on the link scale, it is also the average conditional
link-scale contrast at those means. It is not an outcome-scale marginal
standardization. The model is implemented natively.

```{r}
net_ipd <- cpaic_network(cpaic_bin_agd, ipd = cpaic_bin_ipd, sm = "OR",
                         family = "binomial", ipd_covariates = "x1",
                         inactive = "Placebo")
component_effects(cstc(net_ipd, target = c(x1 = 0),
                       effect_modifiers = "x1",
                       allow_experimental_bridge = TRUE))
```

## Anchored matching-adjusted indirect comparison (cMAIC)

`cmaic()` reweights each IPD study so that its effect-modifier moments match
the supplied target moments, using entropy-balancing weights
$w_i = \exp(\tilde x_i^\top \alpha)$ with $\tilde x_i$ the centered effect
modifiers [@phillippo2020mlnmr]. The effective sample size is
$\mathrm{ESS} = (\sum_i w_i)^2 / \sum_i w_i^2$. The weighted within-study
contrasts, with bootstrap standard errors that propagate the weighting
uncertainty, are then combined through the component model.

That last step is an additional methodology constraint. A marginal weighted
contrast is not generally additive across treatment components on a nonlinear
link scale. Also, if AgD-only edges remain, those contrasts still refer to
their original study populations. cMAIC therefore stops by default outside the
Gaussian identity-link case and whenever retained AgD edges would be mixed in.
The experimental override does not repair either estimand mismatch.

```{r}
fit_maic <- cmaic(net_ipd, target = c(x1 = 0), effect_modifiers = "x1",
                  n_boot = 100, seed = 1,
                  allow_experimental_bridge = TRUE)
effective_sample_size(fit_maic)
```

## The unification

For a single anchored comparison, the contrast algebra resembles a Bucher
indirect comparison. A network bridge combines many adjusted and unadjusted
contrasts through $\hat\beta = (X^\top W X)^{+} X^\top W d$. That algebra does
not establish that every input contrast shares the same target estimand. It
also does not test the required constancy of component effects and
effect-modifier interactions across disconnected sub-networks
[@rucker2021disconnected].

## Component-additive ML-NMR

`cmlnmr()` places the component structure inside multilevel network
meta-regression: the relative effect of an arm is $C\beta$ rather than a
free per-treatment parameter, and aggregate arms are fitted by integrating
the individual model over each study's covariate distribution. Disconnected
sub-networks share the component parameters, so the network is connected by
construction. All four outcome families are supported; survival uses a
proportional-hazards model with a flexible baseline (piecewise-exponential
by default, or a smooth M-spline), set through `cut_points` and `baseline`.
Its analytic row-level likelihood handles events, right/left/interval
censoring, and delayed entry. Aggregate survival arms require reconstructed
event and censoring rows; event counts plus person-time are rejected. The
finite quasi-Monte Carlo covariate integration is approximate.
Aggregate covariates are integrated with a Gaussian copula whose
correlation is estimated from the individual patient data. A component
whose effect-modifier interaction is informed only by aggregate data is
weakly identified and relies on the regression prior.

The default reporting path evaluates
$(C_t-C_u)'(\beta + \Gamma x)$ at a one-row vector of supplied covariate
means. Since this expression is linear in $x$, it is an average conditional
link-scale effect at those means. A separate `marginal_effects()` path accepts
empirical target rows or a named Gaussian-copula target distribution. Empirical
rows must contain exactly the fitted effect modifiers; finite nonnegative
weights are normalized and zero-weight rows are discarded. Bernoulli modifiers
use actual 0/1 rows. Summary targets provide named means, SDs, and margins,
with optional latent-scale correlation and deterministic Sobol' integration.
Continuous modifiers can use normal, gamma, lognormal, or beta margins
independently, while fitted Bernoulli modifiers remain Bernoulli. Independent
Bernoulli strata are enumerated exactly. Summary integration stops when the
requested Bernoulli strata or non-normal moments fail their moment-fidelity
gates, rather than silently reporting a distorted target. The Bernoulli
prevalence tolerance is `max(1e-8, 0.1 * min(p, 1 - p))`; non-normal means must
be within 0.1 target SD and non-normal SDs within 15 percent relative error.
Mixed or non-normal multivariable targets also require resolved pairwise
correlations to agree within 0.05 of a high-resolution deterministic reference
grid.

The estimator averages treatment-specific outcomes within each posterior draw
before forming odds, risk, rate, survival, time-specific hazard, or restricted
mean survival time contrasts. Binomial, rate-difference, and survival measures
require an explicit donor study intercept or baseline hazard. Survival
predictions start at model time zero and are not landmark or delayed-entry
conditioned. Prediction times must remain within the observed follow-up support
of the selected donor study. Survival and risk ratios,
piecewise-exponential RMST ratios, and time-specific marginal hazard ratios are
formed on the log scale with stable log-sum-exp calculations before optional
back-transformation. RMST measures require a piecewise-exponential donor
baseline and are evaluated analytically over its intervals. M-spline fits
support marginal survival, risk, and time-specific marginal hazard measures,
but not RMST. These target checks are not a formal overlap or positivity
diagnostic. Nonlinear marginal effects remain treatment-level
contrasts, not marginal component effects or rankings.

```{r, eval = FALSE}
# Fits with rstan by default; backend = "cmdstanr" is the alternative.
fit <- cmlnmr(ipd, agd, effect_modifiers = "x1", inactive = "Placebo",
              family = "binomial")
component_effects(fit)

# Marginal risk ratio over an empirical target distribution.
marginal_effects(
  fit, target = data.frame(x1 = c(0, 1)), weights = c(0.7, 0.3),
  measure = "risk_ratio", baseline_study = "study_id"
)
```

## Assumptions and caveats

* **Additivity** of component effects. `additivity_test()` can diagnose lack of
  fit where evidence exists, but it cannot validate an unobserved cross-gap
  bridge.
* **Identifiability**: a disconnected network is bridgeable only when
  `cpaic_connectivity()` reports `identifiable = TRUE`.
* **Cross-population transportability** of effect modifiers (the standard
  population-adjustment assumption), here extended to constancy of
  component effects across sub-networks.
* **Two-stage estimand mixing**: retained AgD contrasts remain tied to their
  study populations. The default gate stops this mixed bridge unless the
  caller explicitly opts into research-only behavior.
* **Nonlinearity and non-collapsibility**: MAIC targets a marginal weighted
  contrast and STC a conditional link-scale contrast. Marginal contrasts are
  not generally component-additive on nonlinear links.
* **cML-NMR reporting**: average conditional effects at covariate means and
  marginal effects over target distributions are separate estimands. Marginal
  standardization is treatment-level and does not preserve component additivity
  on nonlinear scales. Measures requiring absolute outcomes transport a
  selected study intercept or baseline hazard. Survival predictions start at
  model time zero, not a landmark or delayed-entry time, and use the donor
  study's observed follow-up support. No formal overlap diagnostic is estimated.

## References
